---
title: "JCSM Revision 1"
author: J. Peter Marquardt
date: "report created on: `r format(Sys.time(), '%Y-%m-%d')`"
knit: (function(inputFile, encoding) {rmarkdown::render(inputFile, output_file = paste0("../Revision_JCSM/", format(Sys.time(), "%Y-%m-%d_%H%M"), "_revision",".html")) } )
output:
  rmdformats::readthedown:
    highlight: kate 
    number_sections: true
    code_folding: hide
    code_download: true
    collapsed: false
---

# Preparing the workspace

```{r, warning=F, message=F}
require(basecamb)
require(dplyr)
require(readr)
require(tableone)

# Read in the data
# use paths from a user-specific paths file (not synchronised with git). This allows for cross-user-functionality
paths <- readr::read_csv('../paths.csv') %>%
  dplyr::pull(path, key)
  
data <- readr::read_csv(paths[['data']], na = '#NULL!') %>%
  basecamb::apply_data_dictionary(data_dictionary = readr::read_csv('Data_dictionary_revision.csv')) %>%
  dplyr::mutate(Insurance_type = dplyr::case_when(Insurance_government == 'Government-sponsored' ~ 'Government-sponsored',
                                                  Insurance_private == 'Private' ~ 'Private',
                                                  Insurance_government == 'Non-government' & Insurance_private == 'Non-private' ~ 'None',
                                                  TRUE ~ 'ERROR_fix_this')
  )

```

# Comment 1: Systematic differences between patients with/without imaging

```{r}
data_tableone <- dplyr::select(data, c('Age', 'Sex', 'Race_ethnicity', 'Marital_status', 'Education_level', 'Insurance_type', 'Cancer_type', 'Months_since_cancer_diagnosis', 'Charlson_comorbidity_index', 'Body_mass_index', 'Muscle_data_present'))

# Comparing patients with muscle data to those without
tableone::CreateTableOne(data=data, strata = 'Muscle_data_present', addOverall = TRUE) %>%
  print(exact = c('Race_ethnicity'), quote = FALSE, noSpaces = TRUE, printToggle = FALSE) -> table1

# Exporting to csv
readr::write_csv(data.frame(table1), 'Table_One_Comparison.csv')

# Creating appealing visual output
tableone::kableone(table1)

```

We find that patients without imaging are older, more likely to be government insured, and more likely to have a longer time from first diagnosis of incurable cancer to index hospital admission.

Differences in cancer type should be taken with a grain of salt due to small subgroups. That said, censored patients are less likely to have lung of gastrointestinal cancer, but more likely to have hematologic malignancies. This is plausible because lung and GI cancer warrant chest or abdominal CT scans required for inclusion. Conversely, staging of leukemia and lymphoma does not necessarily require imaging once an incurable stage is confirmed.

# Comment 2: Median time of follow-up

```{r}
median(data$Time_to_death_LFU_days)
IQR(data$Time_to_death_LFU_days)
```

The median time to death/follow-up was **`r median(data$Time_to_death_LFU_days)` +- `r IQR(data$Time_to_death_LFU_days)` (IQR)** days.
